---
title: "Human Actibity Recognition  model"
subtitle: "Practical Machine Learning, course project"
author: "Michael Mikhailidi"
date: "April 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

# Summary

This paper describes implementation  of the machine learning algorithms and methods to  predict  human activity, using measurements from the wearable devices such as smartphones, smartwatches, ativity trackers and so on. 
The stududy has been done with HAR dataset, collected by the group of the researchers (please check the project page here [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har).
Goal of this work to build mathmatical model, which could identify human activity using data from the wearable devices. 

# Data Preparation 
Fo train and test prediction models we will use two data sets, generously provided by the HAR project contributors. 
Training dataset could be found at [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv).
Testing data are located [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv).

```{r}
set.seed(70720791)
training<- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",header = TRUE)
testing <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",header =TRUE)
```
We will use _training_ data for preparation and mode training, while  _testing_  data set will be used only for the prediction verification.

Structure of the training data are provided below,
```{r}
 dim(training); dim(testing)
summary(training)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
